Emergence of the Southeast Asian islands as a driver for Neogene cooling

### 2020-05-13 Manuscript/Supplementary_Information.tex
PNAS submission checkpoint

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Figures/world_lithology.jpg}
    \caption{Distribution of lithologies at 0.1\degrees $\times$ 0.1\degrees resolution used in GEOCLIM.}
    \label{fig:world_lithology}
\end{figure}

@@

Within the equations that govern the chemical weathering rate in GEOCLIM, we identify the less constrained parameters: the proportionality constant that modifies the dependence of dissolution rate on runoff and temperature ($k_{d}$; Equation 9), the proportionality constant that modifies the dependence of dissolution rate on runoff only ($k_{w}$; Equation 9), the power constant that modifies the dependence of dissolution rate on the time that a rock particle has spent in the regolith ($\sigma$; Equation 4), and the proportionality constant that modifies the dependence of regolith production on runoff and temperature ($k_{rp}$; Equation 6). Furthermore, the Ca+Mg concentrations of metamorphic and siliciclastic sediment grid cells are also difficult to define. We therefore allow these parameters to vary within reasonable bounds during the calibration stage of GEOCLIM. In total, we test 93,600 unique parameter combinations (Fig. \ref{fig:r2_cross_plot}; Table \ref{tab:parameter_combinations}).

@@

We compute spatially-resolved long-term \COtwo consumption (i.e. Ca+Mg fluxes) using present-day runoff (UNH/GRDC Composite Runoff Fields V1.0; \citealp{Fekete1999a}), temperature (CRU TS v.4.03; \citealp{Harris2013a}), and slope (Shuttle Radar Topography Mission; \citealp{Farr2007a}) fields. As described in the main text, we then sum the computed \COtwo consumption over large-scale watersheds that appear in the global compilation of \citet{Gaillardet1999a}, as well as smaller-scale watersheds of the Amazon Basin (HYBAM network) in the compilation of \citet{Moquet2011a}, \citet{Moquet2016a}, and \citet{Moquet2018a}. The latter are nested watersheds, which requires upstream weathering fluxes to be subtracted from downstream fluxes. Watersheds for which this subtraction yields an aberrant value are not considered. 80 watersheds in total are used in this study. We calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these basins:

@@

The majority of the original 93,600 parameter combinations produce \COtwo consumption maps with poor fits to the measured watershed data (Fig. \ref{fig:W_vs_r2}). Given that the coefficient of determination ($r^{2}$) is calculated using Equation \ref{eq:10} rather than fitting a linear model, many of the combinations associated with particularly poor fits result in negative $r^{2}$ values (Fig. \ref{fig:W_vs_r2}). However, given the right permutation of parameters, GEOCLIM produces \COtwo consumption maps that fit the measured watershed data reasonably well. We therefore eliminate all parameter combinations that produce a $r^{2}\leq0.5$, which leaves 573 unique parameter combinations (Figs. \ref{fig:W_vs_r2} and \ref{fig:r2_cross_plot}).

Global CO$_{2}$ consumption calculated from these 573 parameter combinations ranges from 3.7$\times10^{12}$~mol/yr to 6.3$\times10^{12}$~mol/yr, with a mean of 5.2$\times10^{12}$~mol/yr. These estimates fall within the range of those independently estimated in the literature for the present-day: the full range of estimates of the present-day non-anthropogenic global CO$_{2}$ emission rate compiled in \citet{Gerlach2011a} is 3.0--10.0$\times10^{12}$~mol/yr, with the preferred range of those estimates being 3.4--5.9$\times10^{12}$~mol/yr (Fig. \ref{fig:W_vs_r2}B).

Each parameter combination predicts a different total \COtwo consumption that should match the \COtwo degassing at steady-state. Hence, in the non-calibration GEOCLIM experiments, each parameter combination is used with its corresponding steady-state \COtwo degassing. In these estimates we are not including the effects of reverse weathering \citep{Michalopoulos1995a} as it is not clear how it should be parameterized and it is interpreted to be a relatively minor flux in the Cenozoic \citep{Isson2018a}. Formation of authigenic clays that scales with Ca and Mg concentration of riverine waters could decrease the \COtwo consumption associated with silicate weathering although the overall trend seen in this study would remain.

@@

For the climate model component of GEOCLIM, we use temperature and runoff fields from a subset of the GFDL CM2.0 experiments (\citealp{Delworth2006a, Delworth2006b}; available for download at \url{https://nomads.gfdl.noaa.gov/dods-data/gfdl_cm2_0/}) These experiments were performed in order to explore the effect of various changes in forcing agents on climate since ca. 1860 at 2.0\degrees $\times$ 2.5\degrees resolution. In the ``1860 control'' experiment, forcing agents representative of conditions ca. 1860 (including \COtwo, CH$_{4}$, N$_{2}$O, O$_{3}$, sulfates, carbon, dust, sea salt, solar irradiance, and the distribution of land cover types) are held constant for 500 years after reaching equilibrium. \pCOtwo in 1860 is assumed to be 286~ppm. In the ``+1\%/yr to 2$\times$'' experiment, initial conditions are taken from the ``1860 control'' experiment, then \pCOtwo is prescribed to increase from 286~ppm at a compounded rate of +1\% per year for 70 years, when \pCOtwo reaches double (572~ppm) of the initial value. \pCOtwo is then held constant until the end of the 280 year experiment. All non-\COtwo forcing agents are held constant. The ``+1\%/yr to 4$\times$'' experiment is identical to the ``+1\%/yr to 2$\times$'' experiment, except that \pCOtwo is prescribed to increase for 140 years, when \pCOtwo reaches quadruple (1144~ppm) of the initial value. \pCOtwo is then held constant for 160 years. We take the mean of the last 100 years of each of these three experiments (when \pCOtwo is being held constant at its final level) to obtain temperature and runoff fields for 286, 572, and 1144~ppm respectively. Temperature and runoff fields associated with \pCOtwo values between the three modeled \pCOtwo levels (286, 572, and 1144~ppm) are obtained through linear interpolation between the model results.

Given that precipitation in SEAI is primarily driven by large-scale convective upwelling leading to high precipitation over both land and water \citep{Donohoe2017a}, the current approach of using a climate model forced by modern paleogeography for all SEAI scenarios is a reasonable approximation for this region. However, for paleogeographic change that would have very significantly modified precipitation over broad swathes of land such as growth of the Himalaya, this approach of using modern climate model results is not viable. The climate model also assumes the present-day extent of ice-sheets for all \pCOtwo levels. However, without ice-sheets, the global climate would be warmer for a given \pCOtwo (hysteresis loop; \citealp{Pollard2005b}), but the amount of emergent land would be lower. We expect this bias to be minimal, as all of our \pCOtwo estimates fall below the glaciation threshold for Antarctica.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.7\textwidth]{Figures/paleogeographic_reconstructions.pdf}
    \caption{Paleogeographically-reconstructed paleoshorelines for SEAI at 5, 10, and 15~Ma, using the paleogeographic model of \citet{Matthews2016a}.}
    \label{fig:paleogeographic_reconstructions}
\end{figure}

We use geological data to quantify the changes in the area of SEAI over the past 15~m.y. (described in detail in \textit{Paleoshoreline Reconstruction and Geological Synthesis}). To generate the lithologic map used in the ``removed SEAI'' scenario (Fig. \ref{fig:SEAI_fracs}), we remove all land associated with arc-continent collision in SEAI. To generate the lithologic map used in the ``paleo-SEAI'' scenarios (Fig. \ref{fig:SEAI_fracs}), we take the estimated paleoshorelines of SEAI for 15, 10, and 5~Ma, then remove all land that falls outside of these bounds. The slope and lithologic classification of each pixel is identical to that of the present day, provided that these pixels fall within the estimated paleoshorelines of SEAI for their respective time slice. This means that while we are changing the amount of emergent land, we are not changing the latitudinal or longitudinal position and are keeping the slope of emerged pixels the same. Given that similar distributions of lithology and slope are reasonable in the past and that latitudinal translation has been relatively minor, we consider this simplification to be reasonable to first-order (Fig. \ref{fig:paleogeographic_reconstructions}). We then add land in the region that is not exposed today. The assignment of lithology, runoff, and slope for pixels on the Sunda Shelf is not trivial, given that the shelf is not currently exposed. As discussed in \textit{Paleoshoreline Reconstruction and Geological Synthesis}, the Sunda Shelf is a flat-lying, relatively stable platform of continental crust. Islands between Malaysia/Sumatra and Borneo are composed of granite and are surrounded by shallow marine siliciclastic sediments \citep{Darmadi2007a, Hall2009a, Hall2013b}. However, basement highs do not appear in the Gulf of Thailand, and seismic and drill core data suggest siliciclastic fluvial systems draining out toward the South China Sea when the shelf was exposed \citep{Darmadi2007a}. We therefore assign pixels of the Sunda Shelf between Malaysia/Sumatra and Borneo to be 50\% felsic and 50\% sediment, and pixels further to the north to be 100\% sediment. For the slope, we take the mean slope value (0.0043~m/m) of relatively flat land on eastern Sumatra, and assign this value to all pixels of the Sunda Shelf. For the runoff, we linearly interpolate along latitude bands between present-day land pixels to obtain runoff values for the presently-submerged Sunda Shelf. This approach preserves large N-S variation in runoff associated with the large-scale Hadley circulation and therefore should be reasonable to first-order, although it does not accurately capture the smaller-scale interactions between land and ocean/atmosphere circulation. For other minor islands in the region that were exposed 15, 10, or 5~m.y. ago but are currently submerged, we take the same approach as that described for the Sunda Shelf to generate the runoff field. For the slope and lithology, we take the mean slope and Ca+Mg concentrations from neighbouring pixels. We also remove the Greenland ice sheet in the ``paleo-SEAI'' scenarios, to be consistent with Northern Hemisphere ice sheets developing after 5~Ma \citep{Haug2005a}. The runoff field over Greenland is generated using the same method as that for the Sunda Shelf. For the slope, we take the mean slope value (0.0529~m/m) of land for which slope data exists around the edges of Greenland, and assign this value to all pixels of Greenland covered by the ice sheet. We also assign the Ca+Mg concentration of bulk continental crust to these pixels.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Figures/pCO2_proxies.pdf}
    \caption{Proxy-based \pCOtwo estimates for the past 20~m.y. For data that was compiled within \citet{Foster2017a}, error bars indicate standardized uncertainties. For the \citet{Ji2018a} data, error bars represent the 16-84$^{th}$ percentile of resampled estimates. The \citet{Bereiter2015a} data comes from the EPICA Dome C ice core.}
    \label{fig:pCO2_proxies}
\end{figure}

@@

Geological maps, stratigraphic data, and previous paleoshoreline compilations were used to calculate the changes in different types of subaerially exposed rocks in SEAI. Following \citet{Molnar2015a}, we analyzed the area changes of islands that are larger than $\sim$200~km$^{2}$ (Fig. 1 in main text). We also included changes in areas of submerged continental shelves that were previously exposed, like the Sunda Shelf. Larger islands were further divided into regions based upon their position on microplates as defined by \citet{Matthews2016a}. Miocene to present stratigraphic columns were compiled from each region to develop an age model for paleoenvironmental indicators. In general, carbonate strata and thin-bedded siliciclastic strata were assumed to be deposited in a subaqueous marine environment, whereas evidence for exposure and terrestrial siliciclastic deposits were used as indicators of a subaerial paleoenvironment. Following previous paleoshoreline compilations and these environmental indicators, paleoshorelines were outlined in QGIS to calculate areas for the Early Pliocene (5 Ma), Late Miocene (10 Ma), and Middle Miocene (15 Ma). The paleoshoreline reconstructions within the paleogeographic models broadly coincide with sub-Epoch boundaries (i.e. Early Miocene = 23--16~Ma, Middle Miocene = 15.99--11.61~Ma, Late Miocene = 11.6--5.4~Ma, Pliocene =5.39--3.8 Ma) but we recognize that the biostratigraphic resolution is an additional source of uncertainty.

Several new sources for paleoshoreline data have become available since \citet{Molnar2015a}. Particularly, for Sulawesi, we followed the recent stratigraphic compilation and paleoshorelines delineated by \citet{Nugraha2018a}, and in New Guinea, we followed \citet{Gold2017a} and \citet{Harrington2017a}. Although the outlines of our paleoshorelines are significantly different from those used for 5~Ma by \citet{Molnar2015a} our calculated area is comparable.

@@

For the paleoshorelines of Sulawesi, we largely followed the recent stratigraphic compilation and paleoshorelines compilation of \citet{Nugraha2018a}. Geographically, Sulawesi can be divided into a central highland region flanked by North, South, Southeast, and East Arms. These arms have high relief ($>$3~km) that are separated by deep basins and broadly define Sulawesi's seven tectonic provinces: 1) the West Sulawesi magmatic arc of the South Arm, 2) the Central Sulawesi metamorphic Belt, 3) the Sangihe arc of the North Arm, 4) the East Sulawesi Ophiolite of the East Arm, 5) the southeast metamorphic belt of the Southeast Arm, and the microcontinental blocks of 6) Banggai-Sula and 7) Buton-Tukang Besi \citep{Hamilton1979a, Katili1978a}. Like SW Borneo, the North and South Arms and much of Central Sulawesi are underlain by continental blocks that rifted off of the Australian-Birds Head margin in the Jurassic and collided in the Cretaceous with Eurasian basement of Sundaland as part of the Woyla Arc system \citep{Parkinson1998b, Hennig2016a, Hall2017a, Hennig2017a}. After collision, subduction polarity reversed and a Cretaceous to Miocene SE-facing volcanic arc developed \citep{Polve1997a, Elburg1998a} that was connected to the paleo-Sunda Arc \citep{Hall2002b}, also referred to as the Great Indonesian arc \citep{Harris2006a}.  The ophiolites of Sulawesi were generated in this arc system in a back arc to intra-arc setting \citep{Monnier1995a} (but see \citealp{Kadarusman2004a} for an alternative view) and the largest fragments of the East Sulawesi ophiolite were detached from their metamorphic sole in the late Oligocene (32--28~Ma; \citealp{Parkinson1998a}) and thrust to the east above a west dipping slab (East Sulawesi block; \citealp{Villeneuve2001a}). During the Early to Middle Miocene, the Tukang Besi-Buton block began to collide obliquely with the southeastern end of the Sunda Arc, causing ophiolite emplacement in West and SE Sulawesi, including Buton \citep{Smith1991a, Bergman1996a}.

@@

The East and Southeast Arms consist predominantly of mafic and ultramafic rock of the East Sulawesi Ophoilite that is exposed for more than 10,000~km$^{2}$ \citep{Monnier1995a}. The East Arm preserves a complete ophiolitic sequence underlain by a metamorphic sole, m\'elange, imbricate continental margin and crystalline basement with a blueschist metamorphic overprint \citep{Silver1983a, Monnier1995a, Parkinson1998a}. Locally, the structural thickness of the ophiolite exceeds 15~km with surface relief over 3~km \citep{Kadarusman2004a}. Seventeen igneous K-Ar dates from the East Sulawesi ophiolite range from 93--32~Ma, clustering between 60~Ma and 40~Ma \citep{Parkinson1998b}. K-Ar dates on hornblende from the metamorphic sole yielded cooling ages between 36--23~Ma \citep{Parkinson1998a, Villeneuve2001a}, which dates the initial emplacement of the East Sulawesi ophiolite as the north Sulawesi volcanic arc was underthrusted by the Sula spur (Australian crust) \citep{Silver1983a, Parkinson1998a}; however, despite these relatively old ages of emplacement, it appears that the ophiolites on the East Arm were not substantially subaerially exposed and exhumed until the Miocene.

@@

West Sulawesi rifted from Borneo during the Eocene forming the Makassar Straits back arc basin behind a southwest-facing arc \citep{Polve1997a}. Eroded fragments of ophiolite and extensive belts of volcanic rocks are preserved on the West and South Arms of Sulawesi \citep{Bergman1996a, vanLeeuwen2010a}. During the Middle Miocene (ca. 15--13~Ma), extensional faults in the Bone basin reversed, which was accompanied by uplift and erosion of the Bone Mountain ophiolite and Lamasi complex in West Sulawesi \citep{Bergman1996a, vanLeeuwen2010a}. Shortening was likely due to the collision of the leading edge of the Buton-Tukang Besi block, which collided with Buton and the SE Arm of Sulawesi \citep{Smith1991a}. Uplift and erosion is recorded by the presence of a major Middle Miocene unconformity and sedimentary breccias in marginal basins \citep{Bergman1996a, vanLeeuwen2010a}. Uplift was diachronous, not effecting units below and to the west of the Lamasai ophiolite until the Middle Miocene (ca. 13~Ma) \citep{vanLeeuwen2010a}. Alkali volcanism ensued at ca. 11~Ma and is associated with a second phase of extension and exhumation. From the Late Miocene to Pliocene fission track ages imply deep exhumation \citep{Smith1991a, Bergman1996a, vanLeeuwen2010a}.

@@

\section*{Seawater Model}

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Figures/seawater_model.pdf}
    \caption{xxx}
    \label{fig:seawater_model}
\end{figure}

>>>>

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Figures/world_lithology.jpg}
    \caption{Distribution of lithologies at 0.1\degrees $\times$ 0.1\degrees resolution used in GEOCLIM modified from \cite{Hartmann2012a}.}
    \label{fig:world_lithology}
\end{figure}

@@

Within the equations that govern the chemical weathering rate in GEOCLIM, we identify the less constrained parameters: the proportionality constant that modifies the dependence of dissolution rate on runoff and temperature ($k_{d}$; Equation 9), the proportionality constant that modifies the dependence of dissolution rate on runoff only ($k_{w}$; Equation 9), the power constant that modifies the dependence of dissolution rate on the time that a rock particle has spent in the regolith ($\sigma$; Equation 4), and the proportionality constant that modifies the dependence of regolith production on runoff and temperature ($k_{rp}$; Equation 6). Furthermore, the Ca+Mg concentrations of metamorphic and siliciclastic sediment grid cells are difficult to define. We allow these parameters to vary within reasonable bounds during the calibration stage of GEOCLIM. In total, we test 93,600 unique parameter combinations (Fig. \ref{fig:r2_cross_plot}; Table \ref{tab:parameter_combinations}).

@@

We compute spatially-resolved long-term \COtwo consumption (i.e. Ca+Mg fluxes) using present-day runoff (UNH/GRDC Composite Runoff Fields V1.0; \citealp{Fekete1999a}), temperature (CRU TS v.4.03; \citealp{Harris2013a}), and slope (Shuttle Radar Topography Mission; \citealp{Farr2007a}) fields. As described in the main text, we sum the computed \COtwo consumption over large-scale watersheds that appear in the global compilation of \citet{Gaillardet1999a}, as well as smaller-scale watersheds of the Amazon Basin (HYBAM network) in the compilation of \citet{Moquet2011a}, \citet{Moquet2016a}, and \citet{Moquet2018a}. The latter are nested watersheds, which requires upstream weathering fluxes to be subtracted from downstream fluxes. Watersheds for which this subtraction yields an aberrant value are not considered. 80 watersheds in total are used in this study. We calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these basins:

@@

The majority of the original 93,600 parameter combinations produce \COtwo consumption maps with poor fits to the measured watershed data (Fig. \ref{fig:W_vs_r2}). Given that the coefficient of determination ($r^{2}$) is calculated using Equation \ref{eq:10} rather than fitting a linear model, many of the combinations associated with particularly poor fits result in negative $r^{2}$ values (Fig. \ref{fig:W_vs_r2}). However, given the right permutation of parameters, GEOCLIM produces \COtwo consumption maps that fit the measured watershed data reasonably well. We eliminate all parameter combinations that produce a $r^{2}\leq0.5$, which leaves 573 unique parameter combinations (Figs. \ref{fig:W_vs_r2} and \ref{fig:r2_cross_plot}).

Global CO$_{2}$ consumption calculated from these 573 parameter combinations ranges from 3.7$\times10^{12}$~mol/yr to 6.3$\times10^{12}$~mol/yr, with a mean of 5.2$\times10^{12}$~mol/yr. These estimates fall within the range of independently estimated outgassing rates for the present-day: the full range of estimates of the present-day non-anthropogenic global CO$_{2}$ emission rate compiled in \citet{Gerlach2011a} is 3.0--10.0$\times10^{12}$~mol/yr, with the preferred range of those estimates being 3.4--5.9$\times10^{12}$~mol/yr (Fig. \ref{fig:W_vs_r2}B).

Each parameter combination predicts a different total \COtwo consumption that should match the \COtwo degassing at steady-state. Hence, in the non-calibration GEOCLIM experiments, each parameter combination is used with its corresponding steady-state \COtwo degassing. In these estimates, we are not including the effects of reverse weathering \citep{Michalopoulos1995a} as it is not clear how it should be parameterized and it is interpreted to be a relatively minor flux in the Cenozoic \citep{Isson2018a}. Formation of authigenic clays that scales with Ca and Mg concentration of riverine waters could decrease the \COtwo consumption associated with silicate weathering although the overall trend seen in this study would remain.

@@

For the climate model component of GEOCLIM, we use temperature and runoff fields from a subset of the GFDL CM2.0 experiments (\citealp{Delworth2006a, Delworth2006b}; available for download at \url{https://nomads.gfdl.noaa.gov/dods-data/gfdl_cm2_0/}) These experiments were performed in order to explore the effect of various changes in forcing agents on climate since ca. 1860 at 2.0\degrees $\times$ 2.5\degrees resolution. In the ``1860 control'' experiment, forcing agents representative of conditions ca. 1860 (including \COtwo, CH$_{4}$, N$_{2}$O, O$_{3}$, sulfates, carbon, dust, sea salt, solar irradiance, and the distribution of land cover types) are held constant for 500 years after reaching equilibrium. \pCOtwo in 1860 is assumed to be 286~ppm. In the ``+1\%/yr to 2$\times$'' experiment, initial conditions are taken from the ``1860 control'' experiment, then \pCOtwo is prescribed to increase from 286~ppm at a compounded rate of +1\% per year for 70 years, when \pCOtwo reaches double (572~ppm) of the initial value. \pCOtwo is then held constant until the end of the 280 year experiment. All non-\COtwo forcing agents are held constant. The ``+1\%/yr to 4$\times$'' experiment is identical to the ``+1\%/yr to 2$\times$'' experiment, except that \pCOtwo is prescribed to increase for 140 years, when \pCOtwo reaches quadruple (1144~ppm) of the initial value. \pCOtwo is then held constant for 160 years. We take the mean of the last 100 years of each of these three experiments (when \pCOtwo is being held constant at its final level) to obtain temperature and runoff fields for 286, 572, and 1144~ppm \pCOtwo respectively. Temperature and runoff fields associated with \pCOtwo values between the three modeled \pCOtwo levels (286, 572, and 1144~ppm) are obtained through linear interpolation between the model results.

Given that precipitation in the SEAI is primarily driven by large-scale convective upwelling leading to high precipitation over both land and water \citep{Donohoe2017a}, the current approach of using a climate model forced by modern paleogeography for all SEAI scenarios is a reasonable approximation for this region. However, for paleogeographic change that would have very significantly modified precipitation over broad swathes of land such as growth of the Himalaya, this approach of using modern climate model results is not viable. The climate model also assumes the present-day extent of ice-sheets for all \pCOtwo levels. However, without ice-sheets, the global climate would be warmer for a given \pCOtwo (hysteresis loop; \citealp{Pollard2005b}), but the amount of emergent land would be lower. We expect this bias to be minimal, as all of our modeled \pCOtwo estimates fall below the glaciation threshold for Antarctica.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.7\textwidth]{Figures/paleogeographic_reconstructions.pdf}
    \caption{Paleogeographically-reconstructed paleoshorelines for SEAI at 5, 10, and 15~Ma, using the paleoshoreline reconstruction of this study coupled to the paleogeographic model of \citet{Matthews2016a}.}
    \label{fig:paleogeographic_reconstructions}
\end{figure}

We use geological data to quantify the changes in the area of SEAI over the past 15~m.y. (described in detail in \textit{Paleoshoreline Reconstruction and Geological Synthesis}). To generate the lithologic map used in the ``removed SEAI'' scenario (Fig. \ref{fig:SEAI_fracs}), we remove all land associated with arc-continent collision in SEAI. To generate the lithologic map used in the ``paleo-SEAI'' scenarios (Fig. \ref{fig:SEAI_fracs}), we take the estimated paleoshorelines of SEAI for 15, 10, and 5~Ma, then remove all land that falls outside of these bounds. The slope and lithologic classification of each pixel is identical to that of the present day, provided that these pixels fall within the estimated paleoshorelines of SEAI for their respective time slice. This means that while we are changing the amount of emergent land, we are not changing the latitudinal or longitudinal position and are keeping the slope of emerged pixels the same. Given that similar distributions of lithology and slope are reasonable in the past and that latitudinal translation has been relatively minor, we consider this simplification to be reasonable to first-order (Fig. \ref{fig:paleogeographic_reconstructions}). We then add land in the region that is not exposed today. The assignment of lithology, runoff, and slope for pixels on the Sunda Shelf is not trivial, given that the shelf is not currently exposed. As discussed in \textit{Paleoshoreline Reconstruction and Geological Synthesis}, the Sunda Shelf is a flat-lying, relatively stable platform of continental crust. Islands between Malaysia/Sumatra and Borneo are composed of granite and are surrounded by shallow marine siliciclastic sediments \citep{Darmadi2007a, Hall2009a, Hall2013b}. However, basement highs of granite do not appear further to the northwest in the Gulf of Thailand, and seismic and drill core data suggest siliciclastic fluvial systems draining out toward the South China Sea when the Sunda Shelf was exposed \citep{Darmadi2007a}. We therefore assign pixels of the Sunda Shelf between Malaysia/Sumatra and Borneo to be 50\% felsic and 50\% sediment, and pixels further to the north to be 100\% sediment. For the slope, we take the mean slope value (0.0043~m/m) of relatively flat land on eastern Sumatra, and assign this value to all pixels of the Sunda Shelf. For the runoff, we linearly interpolate along latitude bands between present-day land pixels to obtain runoff values for the presently-submerged Sunda Shelf. This approach preserves large N-S variation in runoff associated with the large-scale Hadley circulation and therefore should be reasonable to first-order, although it does not accurately capture the smaller-scale interactions between land and ocean/atmosphere circulation. For other minor islands in the region that were exposed 15, 10, or 5~m.y. ago, but are currently submerged, we take the same approach as that described for the Sunda Shelf to generate the runoff field. For the slope and lithology, we take the mean slope and Ca+Mg concentrations from neighbouring pixels. We also remove the Greenland ice sheet in the ``paleo-SEAI'' scenarios, to be consistent with Northern Hemisphere ice sheets developing after 5~Ma \citep{Haug2005a}. The runoff field over Greenland is generated using the same method as that for the Sunda Shelf. For the slope, we take the mean slope value (0.0529~m/m) of land for which slope data exists around the edges of Greenland, and assign this value to all pixels of Greenland covered by the ice sheet. We also assign the Ca+Mg concentration of bulk continental crust to these pixels.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{Figures/pCO2_proxies.pdf}
    \caption{Proxy-based \pCOtwo estimates for the past 20~m.y. For data that were compiled within \citet{Foster2017a}, error bars indicate standardized uncertainties. For the \citet{Ji2018a} data, error bars represent the 16-84$^{th}$ percentile of resampled estimates. The \citet{Bereiter2015a} data come from the EPICA Dome C ice core.}
    \label{fig:pCO2_proxies}
\end{figure}

@@

Geological maps, stratigraphic data, and previous paleoshoreline compilations were used to calculate the changes in different types of subaerially exposed rocks in SEAI. Following \citet{Molnar2015a}, we analyzed the area changes of islands that are larger than $\sim$200~km$^{2}$ (Fig. 1 in main text). We also included changes in areas of submerged continental shelves that were previously exposed, like the Sunda Shelf. Larger islands were further divided into regions based upon their position on microplates as defined by \citet{Matthews2016a}. Miocene to present stratigraphic columns were compiled from each region to develop an age model for paleoenvironmental indicators. In general, carbonate strata and thin-bedded siliciclastic strata were assumed to be deposited in a subaqueous marine environment, whereas evidence for exposure and terrestrial siliciclastic deposits were used as indicators of a subaerial paleoenvironment. Following previous paleoshoreline compilations and these environmental indicators, paleoshorelines were outlined in QGIS to calculate areas for the Early Pliocene (5 Ma), Late Miocene (10 Ma), and Middle Miocene (15 Ma). The paleoshoreline reconstructions within the paleogeographic models broadly coincide with sub-Epoch boundaries (i.e. Early Miocene = 23--16~Ma, Middle Miocene = 15.99--11.61~Ma, Late Miocene = 11.6--5.4~Ma, Pliocene =5.39--3.8 Ma), but we recognize that the biostratigraphic resolution is an additional source of uncertainty.

Several new sources for paleoshoreline data have become available since \citet{Molnar2015a}. Particularly, for Sulawesi, we follow the recent stratigraphic compilation and paleoshorelines delineated by \citet{Nugraha2018a}, and in New Guinea, we follow \citet{Gold2017a} and \citet{Harrington2017a}. Although the outlines of our paleoshorelines are significantly different from those used for 5~Ma by \citet{Molnar2015a} our calculated area is comparable.

@@

For the paleoshorelines of Sulawesi, we largely followed the recent stratigraphic compilation and paleoshorelines compilation of \citet{Nugraha2018a}. Geographically, Sulawesi can be divided into a central highland region flanked by North, South, Southeast, and East Arms. These arms have high relief ($>$3~km) that are separated by deep basins and broadly define Sulawesi's seven tectonic provinces: 1) the West Sulawesi magmatic arc of the South Arm, 2) the Central Sulawesi metamorphic Belt, 3) the Sangihe arc of the North Arm, 4) the East Sulawesi Ophiolite of the East Arm, 5) the southeast metamorphic belt of the Southeast Arm, and the microcontinental blocks of 6) Banggai-Sula and 7) Buton-Tukang Besi \citep{Hamilton1979a, Katili1978a}. Like SW Borneo, the North and South Arms, and much of Central Sulawesi, are underlain by continental blocks that rifted off of the Australian-Birds Head margin in the Jurassic and collided in the Cretaceous with Eurasian basement of Sundaland as part of the Woyla Arc system \citep{Parkinson1998b, Hennig2016a, Hall2017a, Hennig2017a}. After collision, subduction polarity reversed and a Cretaceous to Miocene SE-facing volcanic arc developed \citep{Polve1997a, Elburg1998a} that was connected to the paleo-Sunda Arc \citep{Hall2002b}, also referred to as the Great Indonesian arc \citep{Harris2006a}.  The ophiolites of Sulawesi were generated in this arc system in a back-arc to intra-arc setting \citep{Monnier1995a} (but see \citealp{Kadarusman2004a} for an alternative view) and the largest fragments of the East Sulawesi ophiolite were detached from their metamorphic sole in the late Oligocene (32--28~Ma; \citealp{Parkinson1998a}) and thrust to the east above a west dipping slab (East Sulawesi block; \citealp{Villeneuve2001a}). During the Early to Middle Miocene, the Tukang Besi-Buton block began to collide obliquely with the southeastern end of the Sunda Arc, causing ophiolite emplacement in West and SE Sulawesi, including Buton \citep{Smith1991a, Bergman1996a}.

@@

The East and Southeast Arms consist predominantly of mafic and ultramafic rocks of the East Sulawesi Ophoilite that are exposed for more than 10,000~km$^{2}$ \citep{Monnier1995a}. The East Arm preserves a complete ophiolitic sequence underlain by a metamorphic sole, m\'elange, imbricate continental margin and crystalline basement with a blueschist metamorphic overprint \citep{Silver1983a, Monnier1995a, Parkinson1998a}. Locally, the structural thickness of the ophiolite exceeds 15~km with surface relief over 3~km \citep{Kadarusman2004a}. Seventeen igneous K-Ar dates from the East Sulawesi ophiolite range from 93--32~Ma, clustering between 60~Ma and 40~Ma \citep{Parkinson1998b}. K-Ar dates on hornblende from the metamorphic sole yielded cooling ages between 36--23~Ma \citep{Parkinson1998a, Villeneuve2001a}, which dates the initial emplacement of the East Sulawesi ophiolite as the north Sulawesi volcanic arc was underthrusted by the Sula spur (Australian crust) \citep{Silver1983a, Parkinson1998a}; however, despite these relatively old ages of emplacement, it appears that the ophiolites on the East Arm were not substantially subaerially exposed and exhumed until the Miocene.

@@

West Sulawesi rifted from Borneo during the Eocene forming the Makassar Straits back arc basin behind a southwest-facing arc \citep{Polve1997a}. Eroded fragments of ophiolite and extensive belts of volcanic rocks are preserved on the West and South Arms of Sulawesi \citep{Bergman1996a, vanLeeuwen2010a}. During the Middle Miocene (ca. 15--13~Ma), extensional faults in the Bone basin reversed, which was accompanied by uplift and erosion of the Bone Mountain ophiolite and Lamasi complex in West Sulawesi \citep{Bergman1996a, vanLeeuwen2010a}. Shortening was likely due to the collision of the leading edge of the Buton-Tukang Besi block, which collided with Buton and the SE Arm of Sulawesi \citep{Smith1991a}. Uplift and erosion is recorded by the presence of a major Middle Miocene unconformity and sedimentary breccias in marginal basins \citep{Bergman1996a, vanLeeuwen2010a}. Uplift was diachronous, not effecting units below and to the west of the Lamasai ophiolite until the Middle Miocene (ca. 13~Ma) \citep{vanLeeuwen2010a}. Alkali volcanism ensued at ca. 11~Ma and is associated with a second phase of extension and exhumation from the Late Miocene to Pliocene with fission track ages implying deep exhumation \citep{Smith1991a, Bergman1996a, vanLeeuwen2010a}.

@@

% \section*{Seawater Model}

% If the weathering of juvenile lithologies of the SEAI played an important role in driving Neogene cooling, the intuitive expectation is that marine \SrSr should have decreased over the Neogene as the global weathering flux of low \SrSr juvenile lithologies increased (compensated by a decrease in the global weathering flux of high \SrSr radiogenic lithologies to maintain constant global \COtwo consumption). Contrary to this expectation, marine \SrSr has been rising since ca. 35~Ma (Fig. \ref{fig:seawater_model}). However, the global weathering flux of isotopically-representative lithologies can be decoupled from marine \SrSr via the regional weathering of isotopically-unique source rocks. For example, the decrease in slope in the marine \SrSr record ca. 15~Ma has been attributed to exhumation of relatively low \SrSr and high \OsOs Outer Lesser Himalaya lithologies \citep{Myrow2015a, Colleps2018a}. This change in slope could also be at least partially driven by the exhumation of mafic and organic-rich forearc sediments in the SEAI during arc-continent collision.

% To test whether the Neogene marine \SrSr record is compatible with an increase in the global weathering flux of low \SrSr juvenile lithologies over the Neogene, we implement a simple seawater model that simulates changes in marine \SrSr that result from changing fluxes from several sources \citep{Brass1975a}:

% \begin{equation*}
%     \frac{dR_{SW}}{dt} = \frac{1}{\tau} \sum_{i} \frac{F_{i}}{F_{tot}}(R_{i} - R_{SW})
% \end{equation*}
